---
title: "HW 01: Mayer's equations and Boscovich's arcs"
subtitle: "Instructions — read these; do the work in the `hw-01.qmd` in your own repo"
execute:
  freeze: auto
  echo: true
  warning: false
  message: false
format:
  html:
    toc: true
    toc-depth: 2
---

::: callout-important
## Due date

This assignment is due **Monday, September 14, at 11:59pm**. To be considered on time:

- Your final rendered document (with all code, output, and answers) pushed to your GitHub
  repo
- That same document, as a PDF, submitted on Gradescope
:::

# Introduction

**Problem 1.** Tobias Mayer had 27 imperfect observational equations but only three unknown
quantities. In 1750, long before least squares became standard, he combined the equations by
sorting them into three deliberately chosen groups of nine, adding within each group, and
solving the resulting three equations.

That grouping rule — sort so the coefficients differ as much as possible between groups —
is **Mayer's invention**, and it is also, honestly, **ad hoc.** Nobody proved it was the
best possible rule; Mayer picked *a* way to combine his equations into 3 groups of 9, not
*the* way. Problem 1 asks: **what happens if you instead
regrouped or reselected his equations at random?**

**Problem 2.** Seven years later, Roger Boscovich faced the same wall on a smaller problem.
He and Christopher Maire had five measurements of the length of one degree of the meridian,
taken at five latitudes from Quito to Lapland, and wanted to know how flattened the Earth is
at the poles — two unknowns, five measurements, and no answer that satisfies all five. 

So Boscovich invented a rule of his own: among a specific subset of lines, take the one that makes the **total absolute error** as small as possible. Problem 2
has you run that rule, then score the same lines by a *different* rule — the sum of squared
errors, which Legendre would publish in 1805 — and see whether the two rules even agree
about which line wins.


## Learning goals

In this assignment, you will...

- develop intuition for how combining observations affects the estimate.
- reproduce historical methods for reconciling measurements that disagree.
- explore how the choice of error criterion changes the answer.

## How this assignment is organized

Everything in this assignment is labelled the same way, in three levels:

| Level | Looks like | What it is |
|---|---|---|
| **Problem** | Problem 1, Problem 2 | One historical episode. There are two |
| **Part** | Part 1a, Part 1b, … Part 2e | One method or one question inside that episode. Lettered |
| **Step** | Step 1, Step 2, … | One numbered thing to do inside a Part |

Your own `hw-01.qmd` uses the **same headings, in the same order**, so Part 2c Step 2 in
these instructions is the section titled "Step 2" under "Part 2c" in your document.

::: callout-important
## How to tell when something has to be handed in

Every Step that needs something from you ends with a line beginning
**"Answer in your `hw-01.qmd`:"**. That line, and only that line, tells you what is being
collected — a filled-in table, a number, or a sentence or two of writing.

Steps without that line are code you run and read; nothing is submitted for them beyond the
output showing up in your rendered document.

In your own `hw-01.qmd`, the same places are marked **"Your answer:"**, followed by blanks,
an empty table, or a line reading *Replace this line with your answer.*
:::

# Getting started

- Go to the [statistical-history](https://github.com/statistical-history) organization on
  GitHub. Click on the repo with the prefix `hw-01`. It contains the starter documents you
  need to complete the assignment.
- Clone it the same way you cloned Lab 01's repo — see
  [Lab 01](https://statistical-history.github.io/labs/lab-01.html) if you
  need the steps again.
- Keep the supplied folder arrangement so the relative paths below resolve correctly.

# Packages and helper functions

**No packages.** Base R only, same as Lab 01.

::: callout-important
## Most of the functions in this assignment are not R functions — we wrote them for you

`solve_mayer_sums()`, `repeat_random_partitions()`, `summarize_alpha_results()` and the
rest are
**custom functions written for this course**. They are not part of R, and they are not in
any package you could install. We wrote them so that this assignment can be about *which
historical method you're running and what it produces*, rather than about writing R code.

Three practical consequences:

| | |
|---|---|
| `?solve_mayer_sums` will not work | Nor will searching the internet for it. It exists only in this course |
| They appear only after `source("homework_01_helpers.R")` runs | If R says `could not find function "..."`, that line didn't run. Re-run it, then try again |
| To see what one actually does, open `homework_01_helpers.R` | Every function is listed in an index at the top of that file, and most have explanatory comments directly above them |

You *will* also use a few **base R functions** — `read.csv()`, `sum()`, `head()`,
`rbind()`, `plot()`, and `lm()`. Those are real R, documented with `?sum` and so on. Each
part's callout says which of the two kinds it is handing you.
:::

All the course-written helpers live in one file, **`homework_01_helpers.R`**.
There is no master list to learn here: **each part below tells you the
helpers that part uses and what each one does for you.**

# Data

We begin by loading the data needed for the first problem. 
**`data/mayer_observations.csv`** is the transcription of Stigler's Table 1.1 (p.22): each row carries a `mayer_group` label (`"I"`, `"II"`,
or `"III"`) recording which of Mayer's three original groups it belongs to, which Part 1a
needs. 

```{r}
#| label: load-data
mayer <- read.csv("data/mayer_observations.csv")

source("homework_01_helpers.R")

check_mayer_data(mayer)
```

`check_mayer_data()` is course-written: it confirms the CSV loaded correctly, stopping with
a readable message if something is wrong and saying nothing at all if it is fine.


You can check what the first couple of rows in the data look like by running `head()` function.

```{r}
#| label: head-data
head(mayer)
```


::: callout-note
## The equation behind Mayer's rows

Each row is one observational equation of the form $\beta - c_i = a_i\alpha +
b_i(\alpha\sin\theta)$, where $\beta$, $\alpha$, and $\alpha\sin\theta$ (shortened to
`gamma_arcminutes` in the data) are the three unknowns Mayer was solving for, and the
constants/coefficients differ row to row because the observations differ. You don't need
the lunar astronomy behind it — just that one row is one equation, and three equations are
enough to produce an answer, though different equations need not agree.

Angles are stored in total arcminutes: $13°10' = 13 \times 60 + 10 = 790$ arcminutes, not
the decimal number 13.10.
:::

# Problem 1 — Mayer's 27 equations

## Part 1a — Reproduce Mayer's hand sums

::: callout-tip
## Helper functions in this part

You do the summing yourself with base R's `sum()`. These three course-written helpers do
the rest:

| Call | What it does for you |
|---|---|
| `format_angle(x)` | `872.83` → `"14 degrees 32.83 minutes"` |
| `solve_mayer_sums(sums)` | Solves your three summed equations for the three unknowns |
| `solve_mayer_rows(mayer, c(9, 16, 19))` | Pulls equations 9, 16, 19 and solves them, in one step |
| `mayer_error_minutes(full, comparison)` | Mayer's 27-vs-3 accuracy comparison |
:::

Mayer's original groups were:

- Group I: equations 1, 2, 3, 6, 9, 10, 11, 12, and 27
- Group II: equations 8, 18, 19, 21, 22, 23, 24, 25, and 26
- Group III: equations 4, 5, 7, 13, 14, 15, 16, 17, and 20

### Step 1 — Split into Mayer's three historical groups

Make three data frames, one per historical group. This starts Group I:

```{r}
#| label: group-one
group_I <- mayer[mayer$mayer_group == "I", ]
```

### Step 2 — Sum each group by hand

For **each** group, use `sum()` to add `constant_arcminutes`, `coef_alpha`, and
`coef_alpha_sin_theta`.
To display a constant sum in degrees and minutes, you may use, for example:

```{r}
#| label: format-angle-example
format_angle(sum(group_I$constant_arcminutes), digits = 0)
```

Hint: Can you replace `sum(group_I$constant_arcminutes)` with anything?

Save each sum under the name its chunk already gives it — `constant_I`, `alpha_I`,
`gamma_I`, then the same for groups II and III. Each is a single number.

**Answer in your `hw-01.qmd`:** fill in this table.

| Mayer group | # equations | Constant sum (arcmin) | Constant sum (deg/min) | Sum of alpha coefs | Sum of alpha-sin-theta coefs |
|---|---:|---:|---:|---:|---:|
| I | | | | | |
| II | | | | | |
| III | | | | | |

### Step 3 — Solve the summed system

The `mayer-sums-table` chunk assembles your nine sums into the `mayer_sums` data frame for
you, so there is nothing to retype. Once they are real numbers rather than `NA`,
uncomment and run:

```{r}
#| label: solve-mayer-sums
#| eval: false
mayer_solution <- solve_mayer_sums(mayer_sums)
mayer_solution
format_angle(mayer_solution$beta_arcminutes)
format_angle(mayer_solution$alpha_arcminutes)
```

::: callout-note
## What `solve_mayer_sums()` returns

$\beta$, $\alpha$, and $\gamma$ are what the 3×3 system solves for, and the system's
constants are in arcminutes — so those three come back in arcminutes too, which is why they
need `format_angle()` to be readable (872.83 is not a number you can picture).

$\theta$ is recovered afterwards from the ratio $\gamma/\alpha$, which
has no units at all, and `solve_mayer_sums()` hands it back as **`theta_degrees`** — already
in decimal degrees. **Report it as it comes.** Do not pass it to
`format_angle()`, which expects arcminutes and would misread it by a factor of 60.

| Unknown | Column | Report it as |
|---|---|---|
| $\beta$ | `beta_arcminutes` | degrees/minutes, via `format_angle()` |
| $\alpha$ | `alpha_arcminutes` | degrees/minutes, via `format_angle()` |
| $\gamma = \alpha\sin\theta$ | `gamma_arcminutes` | arcminutes, as printed |
| $\theta$ | `theta_degrees` | decimal degrees, as printed |
:::

**Answer in your `hw-01.qmd`:** report $\beta$, $\alpha$, $\alpha\sin\theta$, and $\theta$. Recall, we shortened $\alpha\sin\theta$ to be $\gamma$ (gamma).
Your result should land close to Mayer's published $\beta = 14°33'$, $\alpha = 1°30'$,
$\theta \approx -3°45'$ (that is, $-3.75°$ — compare it against your `theta_degrees`
directly) — not identical, since Table 1.1's printed coefficients are rounded.

### Step 4 — Mayer's accuracy claim

Everything in this step is about $\alpha$ alone — set the other two unknowns aside.

**(i) Reproduce his earlier three-equation answer.** Before using all 27, Mayer had
solved only equations 9, 16, and 19 — one from each historical group:

```{r}
#| label: mayer-three-equations
#| eval: false
solve_mayer_rows(mayer, c(9, 16, 19))
```

You now have two answers for $\alpha$ from the same 27 observations: one built from
three equations, one from all of them.

**Answer in your `hw-01.qmd`:** report the $\beta$, $\alpha$, and $\theta$ from the
three-equation solved.

**The argument he then made.** Mayer rounded the two answers: from the three
equations $\alpha$ was about **100** arcminutes; from all 27, about **90**. He wrote each
as the unknown true value $T$ plus whatever error its method carried:

$$100 = T + e_3, \qquad 90 = T + e_{27}$$

$T$ was unknowable — it is the whole reason he was measuring in the first place. But
subtracting one equation from the other removed it, leaving a quantity he could actually
see: $e_3 - e_{27} = 10$. That is one equation in two unknowns, and he closed it with a
claim:

> Because these last values [based on all twenty-seven equations] were derived from nine
> times as many observations, one can therefore conclude that they are nine times more
> correct.
>
> — Mayer (1750), p. 155; quoted in Stigler, *The History of Statistics*, p. 23

That claim is $e_3 = 9\,e_{27}$. Substituting it in:

$$9e_{27} - e_{27} = 10 \quad\Longrightarrow\quad 8\,e_{27} = 10 \quad\Longrightarrow\quad
e_{27} = \frac{10}{9 - 1} = 1.25 \text{ arcminutes}$$

So by his own reasoning his 27-equation figure was good to about a minute and a quarter of arc — a statement about
how wrong he was, reached without ever knowing $T$.

**(ii) Redo these calculations without his rounding.** Run `mayer_error_minutes()` on your own computed
$\alpha$ values (instead of the rounded 90 and 100 arcminutes). Order of arguments matters, so make sure to input the 27-equation $\alpha$ first, then the 3-equation one.

**Answer in your `hw-01.qmd`:** report your `mayer_error_minutes()` value, and one or two sentences on why
it doesn't come out at exactly 1.25.


::: render-commit-push
Commit and push now: *"solved 1a."* 
:::

## Part 1b — Randomly regroup all 27 equations

::: callout-tip
## Helper functions in this part

All three are course-written. `repeat_random_partitions()` is doing the entire repeated
procedure — shuffling, splitting, summing, and solving, a thousand times over — in one call.

| Call | What it does for you |
|---|---|
| `repeat_random_partitions(mayer, repetitions, seed)` | The whole 1,000-repetition procedure. Returns one row per repetition |
| `plot_alpha_runs(results, reference_alpha, main)` | Plots the results in order, with Mayer's own answer marked |
| `summarize_alpha_results(results, reference_alpha, method_name)` | Reduces those results to a single summary row |
:::

::: callout-important
## What the numbers in Parts 1b–1d mean

Read this once; it applies to all three of Parts 1b, 1c, and 1d.

**Everything reported in these three parts is about $\alpha$ alone.** Mayer solved for three
unknowns, but only the alpha coefficient is tracked across repetitions — $\beta$ and
$\alpha\sin\theta$ are set aside for the rest of Problem 1. Every value below is in
**arcminutes**.

**"Smallest", "middle", and "largest" always mean across the 1,000 repetitions.** Each
procedure is run 1,000 times, and each run produces one alpha. The smallest is the lowest
alpha any single one of those 1,000 runs produced; the middle is the median of the 1,000; the
largest is the highest. 

**The "move away" is the distance between one run's alpha and Mayer's own alpha.** For a
single repetition,

$$\text{move} = |\,\alpha_{\text{this repetition}} - \alpha_{\text{Mayer}}\,|$$

where $\alpha_{\text{Mayer}}$ is the value you computed in Part 1a Step 3 from his three
historical groups. It is an absolute distance in arcminutes, so it is never negative and
carries no direction — a run 20′ above Mayer and a run 20′ below both have a move of 20. With
1,000 repetitions you get 1,000 moves, and the summary reports two of them: the middle one and
the biggest one.

**You do not compute any of this by hand.** `summarize_alpha_results()` returns all of it as
one row:

| Column it returns | What it is |
|---|---|
| `repetitions`, `solved` | How many runs were made, and how many produced an answer |
| `smallest_alpha`, `middle_alpha`, `largest_alpha` | The lowest, median, and highest alpha across the 1,000 runs |
| `middle_absolute_move` | The **median** move away from Mayer's alpha |
| `largest_absolute_move` | The **largest** move away from Mayer's alpha — the single most extreme run |
| `theta_not_available` | How many runs produced an alpha but no real $\theta$ |
| `within_5_minutes` | How many runs landed within 5 arcminutes of Mayer's alpha |

In Parts 1c and 1d you also compute a **Mayer-error** with `mayer_error_minutes()`. That is a
different quantity from the move — it is Mayer's own 27-versus-3 comparison from Part 1a Step
4, $|\alpha_3 - \alpha_{27}|/8$ — but it too is about alpha alone, in arcminutes, and its
smallest/middle/largest are likewise taken across the 1,000 repetitions.
:::

Now do 1,000 repetitions where R shuffles the 27 equations, splits them into three
non-overlapping sets of nine, sums each set, and solves — every equation used exactly once
per repetition:

```{r}
#| label: random-partitions
random_partitions <- repeat_random_partitions(
  mayer,
  repetitions = 1000,
  seed = 1750
)
```

### Step 1 — Inspect the results

Look at the first six rows with `head(random_partitions)`.

### Step 2 — Plot the runs in order

Use `plot_alpha_runs()`, passing Mayer's own alpha from Part 1a as the reference.

::: callout-note
## How to read this plot

**Each dot is one repetition.** Its height is the alpha that repetition produced, in
arcminutes, so 1,000 repetitions give up to 1,000 dots.

**The dots are sorted by alpha.** The horizontal axis is *rank*: the
leftmost dot is the smallest alpha of all 1,000 runs, the rightmost is the largest, and the
one in the middle is the median. Sorting is what makes the shape readable: the
curve can only rise from left to right, so a long flat stretch means a great many runs agreed
with each other, and a sharp rise at either end means a few runs went far off.

**The red horizontal line is Mayer's own answer** — the alpha you computed in Part 1a Step 3
from his three historical groups, which is why you pass it in as `reference_alpha`. It is a
fixed value, not a fitted line through the dots.

**The vertical gap between a dot and the red line is that run's move away** (see the callout
above). Dots sitting on the line are runs that reproduced Mayer's answer; the dot furthest
from the line, up or down, is the `largest_absolute_move` your summary reports in Step 4.
:::

### Step 3 — Summarize the runs

Use `summarize_alpha_results()` to reduce the 1,000 runs to a single row. 

### Step 4 — Record your numbers

**Answer in your `hw-01.qmd`:**

- the smallest, middle (median), and largest alpha 
- the largest move away from Mayer's own alpha  — the single most
  extreme of the 1,000 runs,
- how many of the 1,000 runs report `theta_not_available`.


Mayer's 27-versus-3 error arithmetic does not apply here because both his original grouping and every random repartition use all 27 equations. In this part, the movement across alternative groupings is the object you are studying.


::: render-commit-push
Commit and push now: *"solved 1b."* 
:::


## Part 1c — Randomly choose any three equations

::: callout-tip
## Helper functions in this part

`repeat_random_triples()` is the new one; `plot_alpha_runs()` and
`summarize_alpha_results()` are the same two you just used in Part 1b.

| Call | What it does for you |
|---|---|
| `repeat_random_triples(mayer, repetitions, seed, one_from_each_group)` | Draws three equations at random and solves them, as many times as you ask. `one_from_each_group` switches between the two selection rules: when `TRUE`, it selects one equation from each of the three groups. When `FALSE`, it randomly selects three from all twenty seven, with a chance of multiple equations from the same group. |
| `mayer_error_minutes(full, comparison)` | Same helper as Part 1a, now applied to a whole column of results at once |
:::

Instead of using all 27, repeatedly select three distinct equations and solve them
directly:

```{r}
#| label: random-three
#| eval: false
random_three <- repeat_random_triples(
  mayer,
  repetitions = 1000,
  seed = 1751,
  one_from_each_group = FALSE
)
random_three$mayer_error_arcminutes <- mayer_error_minutes(
  mayer_solution$alpha_arcminutes,
  random_three$alpha_arcminutes
)
```

The steps are the same four as Part 1b, on this new set of results.

### Step 1 — Inspect the results

Look at the first six rows with `head(random_three)`.

### Step 2 — Plot the runs in order

`plot_alpha_runs()`, same pattern as Part 1b.

### Step 3 — Summarize the runs

`summarize_alpha_results()`, same pattern as Part 1b.
### Step 4 — Record your numbers

**Answer in your `hw-01.qmd`:** two sets of three — both about alpha, both in arcminutes,
both taken across the 1,000 repetitions:

- the smallest, middle, and largest **alpha**, from the summary row,
- the smallest, middle, and largest **Mayer-error**, which you can read with
  `summary(random_three$mayer_error_arcminutes)`.

Report `largest_absolute_move` and `theta_not_available` too (see Part 1b).


::: render-commit-push
Commit and push now: *"solved 1c."* 
:::


## Part 1d — One equation from each Mayer group

::: callout-tip
## Helper functions in this part

No new ones — this is `repeat_random_triples()` again, with `one_from_each_group = TRUE`
instead of `FALSE`. That single argument is the entire difference between Parts 1c and 1d.
:::

Repeat Part 1c, but require each selection to contain exactly one equation from each of
Group I, II, and III:

```{r}
#| label: one-from-each
#| eval: false
one_from_each <- repeat_random_triples(
  mayer,
  repetitions = 1000,
  seed = 1752,
  one_from_each_group = TRUE
)
one_from_each$mayer_error_arcminutes <- mayer_error_minutes(
  mayer_solution$alpha_arcminutes,
  one_from_each$alpha_arcminutes
)
```

### Step 1 — Inspect the results

Look at the first six rows with `head(one_from_each)`.

### Step 2 — Plot the runs in order

`plot_alpha_runs()`, same pattern as Parts 1b and 1c.

### Step 3 — Summarize the runs

`summarize_alpha_results()`, same pattern as Parts 1b and 1c. Save it as
`one_from_each_summary`.

### Step 4 — Record your numbers

**Answer in your `hw-01.qmd`:** the same six numbers as Part 1c, on this part's results —
all about alpha, all in arcminutes, all across the 1,000 repetitions:

- the smallest, middle, and largest **alpha**, from the summary row,
- the smallest, middle, and largest **Mayer-error**, via
  `summary(one_from_each$mayer_error_arcminutes)`,

plus `largest_absolute_move` and `theta_not_available`.

### Step 5 — Build the comparison table

`rbind()` the three summary rows you saved — `partition_summary` (Part 1b),
`random_three_summary` (Part 1c), and `one_from_each_summary` (Part 1d) — into one
three-row table.

**Answer in your `hw-01.qmd`:** fill in comparison table.

### Step 6 — Compare the two selection rules

**Answer in your `hw-01.qmd`:** two or three sentences comparing this part's
one-from-each-group rule against Part 1c's unrestricted one, using values from your
comparison table.


## Part 1e — Reflection

::: callout-note
**We haven't covered probability yet, so use plain language**, not formal terms. Phrases
like "the answers bunch together," "the answer jumps around," or "this rule seems steadier"
are exactly right. Avoid unexplained formal language like *sampling distribution*,
*standard error*, *confidence interval*, or *statistical significance* — you'll meet the real versions of some of
these ideas later in the course.
:::

**Answer in your `hw-01.qmd`:** one to two paragraphs. In your own words, explain what
happens to the answers under the three alternative procedures — randomly repartitioning all
27 into three sets of nine (Part 1b), choosing any three equations (Part 1c), and choosing
one equation from each Mayer group (Part 1d).

Say which procedure was steadiest and which was most erratic **by a yardstick you name** —
any column of your comparison table will do (`middle_absolute_move`, `largest_absolute_move`,
`within_5_minutes`, the spread between `smallest_alpha` and `largest_alpha`, the
`theta_not_available` count). All of them describe alpha across the 1,000 repetitions. Then
say why Mayer's deliberate grouping may have helped. Different conclusions are fine if they use different named
yardsticks and are backed by your own output — use at least two concrete values from your
results.


::: render-commit-push
Commit and push now: *"Finished problem 1."* 
:::


# Problem 2 — Boscovich's five meridian arcs

Five measurements, two unknowns. Load this problem's data, check it, and add the predictor
the model needs:

```{r}
#| label: arcs-setup
arcs <- read.csv("data/boscovich_arcs.csv")
check_boscovich_data(arcs)

arcs <- add_sin2_latitude(arcs)
arcs
plot_arc_lines(arcs, main = "Five meridian arcs, 1755")
```

Two course-written helpers there: `check_boscovich_data()` validates the file the way
`check_mayer_data()` did for Mayer's, and `add_sin2_latitude()` adds the
$\sin^2(\text{latitude})$ column the model needs. 


Look at that plot before going on. If the Earth were a perfect sphere, these five points
would sit on a flat horizontal line. They don't — but they don't sit on any *single*
sloped line either, which is the whole problem.

**`data/boscovich_arcs.csv`** has five rows, one per measured arc: the `place`, its latitude
in degrees and minutes (`lat_deg`, `lat_min`), and `arc_toises`, the measured length of one
degree of the meridian there. A *toise* is an old French unit, a bit under two metres.

::: callout-note
## The equation behind Boscovich's rows, if you're curious

Each row is one observational equation of the form $y_i = A + B\sin^2
\theta_i$, where $y_i$ is the measured arc length, $\theta_i$ is the latitude, and $A$ and
$B$ are the two unknowns. If the Earth is flattened at the poles, a degree of latitude gets
*longer* as you go north, and it does so in proportion to $\sin^2(\text{latitude})$ — that's
the whole physical content of the model. $A$ is the arc length at the equator; $B$ says how
fast it grows.

The quantity Boscovich actually reported is the **ellipticity**, $3A/B$ — the helpers give
it as `one_over_ellipticity`. A value of 230 means the Earth's polar radius is about
$1/230$ shorter than its equatorial radius. A *negative* value means the line slopes the
wrong way, i.e. an Earth stretched at the poles rather than squashed.
:::

Every method in this problem gets judged by the same scorecard, `score_line()`, which
reports four things about any line you hand it:

| Column | What it means |
|---|---|
| `sum_abs_residuals` | total absolute error — **Boscovich's criterion** |
| `sum_squared_residuals` | total squared error — **Legendre's criterion**, from 1805 |
| `residual_total` | the errors added up *with* their signs — Boscovich's requirement that they cancel to zero |
| `one_over_ellipticity` | $3A/B$, the shape of the Earth this line is claiming |

## Part 2a — Boscovich's method of situation

::: callout-tip
## Helper functions in this part

The mathematical fact above is what makes the method computable by hand. The helper does
that reduction for you:

| Call | What it does for you |
|---|---|
| `boscovich_candidate_lines(arcs)` | Builds all five candidate lines *and* scores each one against all five arcs (points). One call, the whole method |
| `boscovich_residual_table(arcs)` | Lays out all 25 residuals as a table, so you can see each candidate line measured against each arc (point)|
| `best_line(lines, criterion = "absolute")` | Picks the winner under Boscovich's own criterion |
| `plot_arc_lines(arcs, lines, main)` | Draws the arcs with any set of lines on top |
:::

Boscovich wanted the line that makes the total absolute error as small as possible, with the
extra requirement that the errors **cancel out to zero**. That second requirement forces the
line through the centre of the data, $(\bar{x}, \bar{y})$ — which kills one of the two
unknowns, leaving only the slope free. And then a genuine mathematical fact about this
problem takes over: **the best such line always joins the centre to one of the five measured
arcs.** So there are only five lines to check.

That reduction is what made the method usable by hand in 1755. Run it:

```{r}
#| label: boscovich-candidates
candidates <- boscovich_candidate_lines(arcs)
candidates
```

### Step 1 — Read all 25 residuals

Each of the five candidate lines is scored against **all five** arcs, not just the one
that defined it. See all 25 residuals at once:

```{r}
#| label: boscovich-residual-table
boscovich_residual_table(arcs)
```

**Every candidate has a residual of exactly zero at its own anchor.**

**Answer in your `hw-01.qmd`:** one sentence explaining why that has to be true, and why it
does *not* mean that candidate is a good line.

### Step 2 — Pick Boscovich's answer

The candidate with the smallest total absolute error:

```{r}
#| label: boscovich-best
#| eval: false
boscovich_answer <- best_line(candidates, criterion = "absolute")
boscovich_answer
```

### Step 3 — Fill in the candidate table

**Answer in your `hw-01.qmd`:** fill in this table from `candidates`.

| Candidate line (centroid + …) | Slope $B$ | Sum \|resid\| | 1/ellipticity |
|---|---:|---:|---:|
| Quito | | | |
| Cape of Good Hope | | | |
| Rome | | | |
| Paris | | | |
| Lapland | | | |

### Step 4 — Plot the candidates

Plot all five, then the winner alone:

```{r}
#| label: plot-candidates
#| eval: false
plot_arc_lines(arcs, candidates, main = "Boscovich's five candidate lines")
plot_arc_lines(arcs, boscovich_answer, main = "Boscovich's answer")
```

Now read your `one_over_ellipticity` column against what people thought was real Earth's value of about **230**.

**Answer in your `hw-01.qmd`:** A negative value and a wildly large one mean different
things — say what each one means.

## Part 2b — The same five lines, scored by squares

::: callout-tip
## Helper functions in this part

**No new lines** — you reuse the `candidates` table you already built and
call `best_line()` on it again, switching `criterion = "absolute"` to
`criterion = "squared"`. That one argument is the whole of Legendre's disagreement with
Boscovich.
:::

Fifty years later Legendre proposed a different scoreboard: minimize the sum of *squared*
errors instead of absolute ones. He gave no argument that squares were *right*, only that
the resulting arithmetic was tractable.

### Step 1 — Rescore the same five candidates

**No new lines are needed.** The same five candidates get rescored:

```{r}
#| label: boscovich-squares
#| eval: false
best_line(candidates, criterion = "squared")
```

**Answer in your `hw-01.qmd`:** does the winner change? Report both winners and their
scores.

### Step 2 — Fill in the second scoreboard

**Answer in your `hw-01.qmd`:** fill in this table, next to the one from Part 2a Step 3.

| Candidate | Sum \|resid\| | Sum resid² |
|---|---:|---:|
| Quito | | |
| Cape of Good Hope | | |
| Rome | | |
| Paris | | |
| Lapland | | |

That is all Part 2b asks. Both error criteria agree about the winner here — Part 2c is
where they stop agreeing, and where you work out why.

## Part 2c — Every pair of arcs

::: callout-tip
## Helper functions in this part

| Call | What it does for you |
|---|---|
| `all_pair_lines(arcs)` | Builds and scores all ten two-arc lines in one call |
| `best_line(lines, criterion)` | The same helper from Parts 2a–2b, now pointed at the pairs table instead of the candidates table |
:::

Before 1750 the standard practice was simpler: pick your two best measurements and ignore
the rest. Any two arcs fix $A$ and $B$ exactly. With five arcs there are only
$\binom{5}{2} = 10$ such pairs, so you can just compute every one of them — no sampling
needed, unlike Problem 1's 27 equations.

```{r}
#| label: all-pairs
#| eval: false
pair_lines <- all_pair_lines(arcs)
pair_lines
```

### Step 1 — The range across the ten pairs

**Answer in your `hw-01.qmd`:** the range of `one_over_ellipticity` across all ten pairs,
and how many pairs give a *negative* value.

### Step 2 — Do the two scoreboards rank the ten the same way?

You now have both scoreboards for the same ten lines.

**Answer in your `hw-01.qmd`:** find a pair of lines where absolute error prefers one and
squared error prefers the other, and report both lines' scores under both criteria. Then,
in two or three sentences: why might that be the case?

### Step 3 — Best pair under each criterion

```{r}
#| label: best-pairs
#| eval: false
best_line(pair_lines, criterion = "absolute")
best_line(pair_lines, criterion = "squared")
```

Now compare the best pair against Boscovich's answer from Part 2a, under **both**
criteria. You should find that the best pair **beats** Boscovich under one criterion and
**loses** to him under the other.

**Answer in your `hw-01.qmd`:** all four numbers, in the table provided.

::: callout-note
## That is not the contradiction it looks like

Boscovich never allowed himself every line. His first requirement was that the errors
**cancel to zero**, and all five of his candidates satisfy it — look at the
`residual_total` column for them, and then at the same column for the pair lines. The
winning pair does not cancel; its errors lean systematically one way. It buys its lower
absolute total with a move Boscovich had ruled out, so it was never in his competition
to begin with.
:::


## Part 2d — Mayer's method on Boscovich's five equations

::: callout-tip
## Helper functions in this part

| Call | What it does for you |
|---|---|
| `mayer_split_line(arcs, low_group_size)` | Sorts the arcs, cuts them into a low group and a high group at the size you name, averages each group, solves the resulting 2×2, and scores the answer — Mayer's whole method in one call |
:::

Something that we can think of as Mayer's rule from Problem 1 wasn't about lunar astronomy — it was a general recipe for
too-many-equations problems: sort the equations so the coefficients differ as much as
possible between groups, combine within each group, and solve. Boscovich's problem has five
equations and **two** unknowns, so it needs **two** groups, not three.

Sort the arcs by $\sin^2(\text{latitude})$ and cut them into a low group and a high group.
There are two sensible cuts:

```{r}
#| label: mayer-splits
split_2_3 <- mayer_split_line(arcs, low_group_size = 2)
split_3_2 <- mayer_split_line(arcs, low_group_size = 3)

rbind(split_2_3, split_3_2)
```

### Step 1 — Report both splits

**Answer in your `hw-01.qmd`:** each split's two groups, its `coefficient_spread` (how far
apart the two group averages are), and both of its residual scores.

### Step 2 — Spread vs. fit

**Mayer's own stated rule says to maximize the spread between groups.**

**Answer in your `hw-01.qmd`:** which split has the larger `coefficient_spread`? Which
split actually fits better? Report the two spread values and the two
`sum_squared_residuals` values that settle it.

### Step 3 — Plot the better split

Take the split with the **lower** residual scores — the one you picked out in Step 2 —
and draw its line against the five arcs:

```{r}
#| label: plot-better-split
#| eval: false
better_split <- if (split_2_3$sum_abs_residuals < split_3_2$sum_abs_residuals) {
  split_2_3
} else {
  split_3_2
}

plot_arc_lines(arcs, better_split, main = "Mayer's line on the five arcs")
```

**Answer in your `hw-01.qmd`:** where does that line fall relative to the five measured
points themselves? Mayer's method has quietly done something neither Boscovich's five
candidates nor the ten pairs could do — say what, in one sentence.

## Part 2e — Lines that miss every point

::: callout-tip
## Helper functions in this part

| Call | What it does for you |
|---|---|
| `score_line(arcs, intercept, slope, label)` | Scores *any* line you hand it. Here you hand it `lm()`'s answer, so least squares gets judged on exactly the same scorecard as every hand method |
| `compare_methods(...)` | Stacks scored lines from all four methods into one table |

`lm()` is **real base R** functio —`?lm` works. Lab 01 introduced it, but we treat it as a black box for now.
:::

::: callout-warning
## Do Step 1 before you run any more code
:::

### Step 1 — Predict, before running any more code

Every method so far searched a short list you could write down: five candidate lines
(Part 2a), ten pair lines (Part 2c), one line per split (Part 2d).

**Answer in your `hw-01.qmd`:** two or three sentences, in plain language, answering: is
there any reason to think the best line is on one of those lists at all? If you wanted to
check a line that isn't — say, a slope halfway between two of your candidates — how would
you go about deciding whether it was better, and what would stop you from doing that for
every possible slope by hand in 1755?

*Write your answer before running Step 2. It is graded on being a real prediction, not on
being right.*

### Step 2 — Least squares

Now use the tool Lab 01 introduced. Treat it as a black box for now — what it does
internally is a Chapter 4 question:

```{r}
#| label: least-squares
#| eval: false
fit <- lm(arc_toises ~ sin2_latitude, data = arcs)
coef(fit)

least_squares <- score_line(
  arcs,
  intercept = coef(fit)[1],
  slope = coef(fit)[2],
  label = "least squares (lm)"
)
least_squares
```

`lm()` does not pick from a list. It searches **every** line there is and returns the one
with the smallest `sum_squared_residuals` — the exact quantity you have been computing by
hand all along.

**Answer in your `hw-01.qmd`:** one or two sentences commenting on the results you see.
Where does `lm()`'s line land next to the ones you built by hand, and what was `lm()` free
to do that none of those methods were?


::: render-commit-push
Commit and push once both problems are done, with a message describing what's in the
commit.
:::

# Submission

::: callout-warning
Before you wrap up, make sure your Git pane is empty — everything committed and pushed.

You must also **render your document to PDF and submit that PDF on Gradescope** before the
deadline for full credit.
:::

To submit on Gradescope:

- Access Gradescope through the menu on the
  [STA 119FS Canvas site](https://canvas.duke.edu/courses/84980).
- Click on the assignment, and you'll be prompted to submit it.
- Mark the pages associated with each exercise — every page of your submission should be
  associated with at least one question.
