---
title: "HW 02: How much evidence is enough?"
subtitle: "Instructions — read these; do the work in the `hw-02.qmd` in your own repo"
execute:
  freeze: auto
  echo: true
  warning: false
  message: false
format:
  html:
    toc: true
    toc-depth: 2
---

::: callout-important
## Due date

This assignment is due **Wednesday, September 23, at 11:59pm**. To be considered on time:

- Your final rendered document (with all code, output, and answers) pushed to your GitHub
  repo
- That same document, as a PDF, submitted on Gradescope
:::

# Introduction

HW 01 gave you two ways to combine observations that disagree — Mayer's grouping and
Boscovich's method of situation. Both produced a number. Neither produced a *reason to
trust the number*. Mayer's own attempt at one ("nine times as many observations, therefore
nine times more correct") was an assertion, not a derivation.

Jacob Bernoulli went at that gap from the other direction. Instead of starting with
observations and hunting for a rule, he started with a mechanism whose chances he *already
knew* — an urn with a stated number of red and black balls — and asked: **how many draws
before the observed frequency is dependably close to the true proportion?** "Dependably" had
to be made precise before the question had an answer, and making it precise is most of the
work.

In the example from lecture, the true proportion is $30/50$, the allowed gap is $1/50$, and
the required odds are $1000:1$ in favor of landing inside that gap. Bernoulli's answer was
25,550 draws. De Moivre's later work made probabilities for large collections of trials far
easier to approximate, which mattered enormously when every such number had to be computed
by hand.

This assignment has three problems:

| Problem | What you do | What it is for |
|---|---|---|
| **1** | Probability calculations on a known urn | The model has to exist before it can be simulated |
| **2** | Build that urn in R and rebuild Bernoulli's question | Find an $N$ by simulation, and compare it to his bound |
| **3** | Design the same standard for two games of different rarity | Then try to break both, and say what passing really proved |

## Learning goals

In this assignment, you will...

- work out probabilities, expectations, and variances from a fully specified model.
- see what averaging does to variation, and why counts and proportions behave differently.
- distinguish the size of one experiment from the number of experiments.
- decide what evidence would make you call a random mechanism honest — and find the limits
  of that decision.
- see what changes when the same standard is applied to a common event and a rare one.

## Reading

Stigler, **pp. 62–77** (Bernoulli and the expanded binomial) and **pp. 78–87** (de Moivre's
approximation and its limits), with a bridge to **pp. 88–98** (Simpson, Bayes, and
assumptions about errors).

Parts 1b and 1c adapt Exercises 2.2 and 2.7 of Peter Hoff's *A First Course in Bayesian
Statistical Methods* (2009), pp. 225–226. **Everything you need is stated below — you do not
need another book.** If you want extra review of random variables, expectations, and
variances, [Hoff's text](https://sites.math.rutgers.edu/~zeilberg/EM20/Hoff.pdf) §2.4 is
freely available.


Your own `hw-02.qmd` uses the **same headings, in the same order** as these instructions, so Part 2c Question 4 in
these instructions is item 4 under the heading "Part 2c" in your document.

::: callout-important
## Caution
**This assignment also collects code.** Several chunks in your template are deliberately
unfinished and carry `#| eval: false` so they display without running. Replace the `TO DO`
comments, the `NA` values, and any `stop("Complete ...")` line, then **delete the
`#| eval: false` line** so the chunk actually runs.
:::

# Getting started

- Go to the [statistical-history](https://github.com/statistical-history) organization on
  GitHub. Click on the repo with the prefix `hw-02`. It contains the starter documents you
  need.
- Clone it the same way you cloned HW 01's repo — see
  [Lab 01](https://statistical-history.github.io/labs/lab-01.html) if you need the steps
  again.
- Keep the supplied folder arrangement so `source("homework_02_helpers.R")` resolves.
- Run the document top to bottom before you submit, and **keep the seeds you used** so your
  results can be reproduced.

# Packages and helper functions

**No packages.** Base R only, same as HW 01.


# Problem 1 — What follows from a probability model?

Show your reasoning for every answer. You may use R for arithmetic, but **a number from
a simulation does not replace a probability calculation here.**

::: callout-note
## Working by hand

You may do any part of this problem on paper or a tablet. If you do:

- Label each part, so a grader can tell 1a from 1b without guessing.
- **Merge** the scan into your rendered PDF and submit **one file**, not two.
- Say so in your `hw-02.qmd` — a line under the part, naming where the work is.
- Page-match the handwritten pages in Gradescope along with the rest.
:::

## Part 1a — The contents of the urn

An urn contains **30 red balls and 20 black balls**. Each ball is equally likely to be
selected. After each draw, the ball is returned and the urn is mixed thoroughly.

1. Find the probability of red on one draw, red on **both** of two draws, and **at least
   one** red in two draws. Explain where you use independence.
2. Let $X$ count red balls in five draws. Find $P(X=3)$, explaining the counting factor in
   your calculation.
3. Now draw twice **without replacement**. Find the probability of two reds. Why did it
   change?
4. Return to sampling with replacement. Could observing only the colors, even for a very
   long time, distinguish this urn from one containing **three red and two black** balls?
   What can be learned about the contents, and what remains unidentified?

**Answer in your `hw-02.qmd`:** the calculations and a short explanation for each question.

## Part 1b — What does averaging buy us?

*Adapted from Hoff, Exercise 2.2.*

1. For independent $Y_1, Y_2$ with means $\mu_1, \mu_2$ and variances $\sigma_1^2, \sigma_2^2$, 
derive the mean and variance of $aY_1 + bY_2$ and of $aY_1 - bY_2$. **State where
independence is needed** — one of the two derivations needs it and the other does not. 
You can use linearity of expectation formula. For variance derivations, use the definition of variance.
In class we have seen that for general $X, Y$, $Var(X+Y) = Var(X)+Var(Y)+2Cov(X,Y)$. Do not just apply this formula.

Then apply your results to independent red-draw indicators with $p = 0.6$:

2. Find the mean and variance of $(Y_1 + Y_2)/2$ (using part 1).
3. Now replace $Y_2$ by a **copy of $Y_1$** — the same observation written down twice. Does
   averaging still reduce variation?
4. For $N$ independent draws, give the mean and standard deviation of both the red count
   $X_N = Y_1 + \dots + Y_N$ and the proportion $\widehat p_N = X_N/N$. If $N$ is multiplied by **four**, what
   happens to each standard deviation? Connect this distinction to the scale of fluctuations
   in de Moivre's work (revisit page 83 and table 2.1 in Stigler).

**Answer in your `hw-02.qmd`:** the derivations, the numerical two-draw comparison, and the
count-versus-proportion explanation.

## Part 1c — Probability as a fair price

*Adapted from Hoff, Exercise 2.7.*

You and a friend are betting on **one draw** from our urn: 30 red, 20 black.

Here is the bet. A **red ticket** pays 100 cents if the ball is red and nothing if it is
black. **You choose the ticket price.** Your friend then decides which of these two ways you
will play:

| Your friend's choice | What happens now | What happens after the draw |
|---|---|---|
| **You buy** | You pay your friend the price | If red, your friend pays you 100 cents |
| **You sell** | Your friend pays you the price | If red, you pay your friend 100 cents |

All of this is agreed **before** the ball is drawn. Count both payments when working out
what you make or lose, and use cents throughout.

1. **What would be a fair price?** Suppose you *buy* the ticket. What price would make your
   expected profit zero? Show the calculation. Would that same price give you zero expected
   profit if you *sold* instead? Could you still lose money on a single draw?

2. **Can your friend make you lose either way the draw turns out?** Suppose you set the
   price at **110 cents**. Would your friend choose to buy from you or sell to you? Work out
   how much you lose if the ball is red and if it is black.

   Now try a price of **−10 cents**. A negative price means the seller pays the buyer 10
   cents to take the ticket, and still owes the buyer 100 cents if red comes up. What would
   your friend choose this time?

   **What range of prices avoids these guaranteed losses?**

3. **What if you bet on both colors?** A **black ticket** pays 100 cents if the same draw is
   black and nothing if it is red.

   You price the red ticket at **60** cents and the black ticket at **50** cents. Your
   friend sells you one of each. How much do you pay? How much do you receive after the
   draw? How much do you make or lose?

   Now lower the black ticket to **30** cents. Your friend buys both tickets from you
   instead. Work out what you receive and what you must pay out.

   **Why must the red and black prices add to 100 cents** to avoid giving your friend a way
   to win every time? What does this example tell you about probability? 

**Answer in your `hw-02.qmd`:** your calculations for each question, and a short explanation
of what the betting example tells you about probability.

::: render-commit-push
Commit and push now: *"finished problem 1."*
:::

# Problem 2 — Reconstruct Bernoulli's question in R


::: callout-important
## The helpers summarize your experiments. They do not run them for you.

`summarize_experiments()`, `simulate_grid()`, `plot_certainty()` and `check_indicators()`
are **course-written functions**, in `homework_02_helpers.R`. They are not part of R and not
in any package, so `?simulate_grid` will not work.
Open the file — there is an index at the top and comments above each function.

**You write the three simulators yourself:** `draw_urn()` in Part 2a, and `play_heads()` and
`play_flush()` in Part 3b. The point of those parts is to practice writing functions.

| Call | What it does for you |
|---|---|
| `summarize_experiments(counts, N, p0, epsilon, certainty_odds)` | Takes success counts from many experiments; reports how often the proportion landed inside the band |
| `simulate_grid(N_values, p0, epsilon, certainty_odds, B, model_p)` | Runs a whole binomial study at each $N$ you name, and stacks the summaries |
| `plot_certainty(results, certainty_odds)` | Plots estimated $q_N$ against $N$ with the target marked |
| `check_indicators(x, N)` | Confirms a vector is exactly $N$ zeros and ones |

You will also use **real base R**: `sample()`, `replicate()`, `cumsum()`, `seq_along()`,
`choose()`, `rbinom()`, `pnorm()`, `pbinom()`. Those are documented — `?sample` works.
:::


Every helper table uses the same columns:

| Field | Meaning |
|---|---|
| `N` | Trials in **one** experiment |
| `B` | Number of independent experiments simulated |
| `inside`, `outside` | How many experiments landed in / out of the band |
| `q_hat` | Estimated probability of landing inside — `inside / B` |
| `odds_hat` | Estimated inside/outside odds. `NA` means **no** outside cases were seen |
| `lower95`, `upper95` | Monte Carlo uncertainty interval for $q_N$ |
| `assessment` | That interval compared against your certainty target |

## Part 2a — Build an urn and watch the proportions

1. Write `draw_urn(N)` so that it returns a vector of `N` zeros and ones, with **one meaning
red**. Represent the urn as 30 ones and 20 zeros, and use `sample()` to draw **with
replacement**. Your template gives the function shell and a syntax example.

```{r}
#| label: sample-syntax-example
sample(c("small", "medium", "large"), size = 8, replace = TRUE)
```

Then:

2. Generate **three independent sequences**, each with 25,550 draws. Set a seed **once,
   before** generating them — never inside the function, or all three sequences come out
   identical.
3. For each sequence, calculate all the running proportions with
   `cumsum(draws) / seq_along(draws)`.
4. Plot the three paths together. Add horizontal lines at $0.60$, $0.58$, and $0.62$. A
   logarithmic horizontal axis (`log = "x"`) makes the early draws visible; the template
   already does this.
5. Identify the **independent unit** being simulated. What role does returning the ball
   after each draw play in keeping the red probability fixed and the draws independent
   under this model?
6. Inspect the paths and answer two questions. **Must the distance from $0.60$ decrease at
   every draw?** **Does first entering the band mean a path will stay there forever?**

::: callout-tip
## Read the plot before you answer

Question 5 is asking about the *shape* of the paths, not about the endpoint. Look at where
each path first crosses into the band, and then look at whether it is still inside a few
hundred draws later.
:::

**Answer in your `hw-02.qmd`:** your simulator, the plot, and your answers to Questions 4
and 5.

## Part 2b — One long sequence is not a probability

One sequence of 25,550 draws gives you **one** proportion. A probability is a statement
about how often something happens across *repetitions of the whole experiment*, so we need
two different counts:

| Symbol | Meaning |
|---|---|
| $N$ | Draws in **one experiment**; its outcome is one proportion $\widehat p_N$ |
| $B$ | Independent repetitions of **the whole experiment**; these estimate how often the proportion falls inside the band |

Let

$$q_N = P\left(\left|\widehat p_N - p_0\right| \leq \epsilon\right).$$

Here $p_0 = 0.6$ and $\epsilon = 1/50$, and **the boundary is included** — a proportion of
exactly $0.58$ counts as inside. Our operational version of Bernoulli's certainty
requirement is

$$\frac{q_N}{1 - q_N} \geq c \quad\Longleftrightarrow\quad q_N \geq \frac{c}{c+1}.$$

::: callout-warning
## What $c = 1000$ is and is not
$c:1$ are **odds in favor of landing inside the band**, and the corresponding probability is $\frac{c}{c+1}$.
$1000/1001 \approx 0.999001$ is a probability corresponding to odds of 1000:1. It is not a
probability of 1000, it is not the ratio of red balls to black balls, and it is not the
probability that the urn is honest. Do not round it to 0.999 — the whole exercise in Part 2c
lives in the gap between those two numbers.
:::

1. Set $N = 100$ (sample size), $B = 2000$ (how many times we repeat the experiment), $p_0 = 0.6$ (true proportion of reds) and $\epsilon = 1/50$ ("grace" window).
Use your `draw_urn()` to repeat the whole experiment, saving its red count each time — a loop or `replicate()`
   will do it, and the template supplies the surrounding code. Recall, `draw_urn()` returns a vector of length $N$ of 1's (reds) and 0's (blacks), not a count. Your `counts` should have
   **$B$ entries, each between 0 and $N$**; it is not one long vector of individual draws.
2. Run `summarize_experiments()` and display its table. Report $N$, $B$, `inside`,
   `outside`, and `q_hat`.
3. Calculate the inside/outside ratio yourself from the two counts and compare it with
   `odds_hat`. If the outside count is zero, report **zero observed failures** — say so in
   those words, and do not read it as infinite certainty.
4. Explain **why increasing $B$ answers a different question from increasing $N$.** Name
   which one changes the size of each experiment, and which one gives you more repetitions
   for estimating its chance of landing inside the band.

**Answer in your `hw-02.qmd`:** your repetition code; $N$, $B$, the two counts, the estimated
$q_N$, and the inside/outside ratio; and your explanation of $N$ versus $B$.

## Part 2c — How large does N need to be?

For large studies, `rbinom(B, size = N, prob = p0)` simulates $B$ red counts directly. Under
the assumptions you checked in Part 2a this has exactly the same count distribution as your
urn, and it avoids constructing billions of individual draws. `simulate_grid()` (in `homework_02_helpers.R`) uses this
shortcut.

Part 2b estimated $q_N$ at exactly one experiment size, $N = 100$. Now you repeat that
study at several sizes and ask when the probability of landing in $[0.58, 0.62]$ reaches
the target. **Keep the urn, the band and the target fixed; the only thing that changes is
the number of draws in one experiment.** There are three stages — a **pilot** at widely
spaced sizes, a **refinement** near the apparent transition, and a **confirmation** at two
sizes you pick from the search. You are not expected to guess a good $N$ in advance.

1. **Run the pilot** with $B = 10000$ at
   `N_values = c(100, 500, 1000, 2500, 5000, 7500, 10000, 25550)`, using $p_0 = 0.6$,
   $\epsilon = 1/50$, and $c = 1000$. Keep every row, including the sizes that fall short.
2. **Read it.** Display the results and use `plot_certainty()` to plot $\widehat q_N$
   against $N$ with the target marked. Going down the rows in increasing $N$, where do the
   estimates move from below the target to above it? Check `assessment` too — a point above
   the line can still be `unresolved` if its interval crosses the line (`lower95` is below target probability). **Do not round
   `q_hat` to 1 before making that comparison**; the whole question lives in the fourth
   decimal place.
3. **Refine.** Add at least four values of $N$ between the last pilot value below the target
   and the first one above it. If the pilot is unclear, cover the unresolved region instead.
   **Keep every value you tried** — do not replace earlier results with whichever run looks
   best.
4. **Choose before you confirm.** Pick **one candidate $N$ and one smaller comparison $N$**
   from your search, and write down why those two, before running anything else. A
   promising candidate is one whose results suggest it could meet the target; the smaller
   size is what shows you what your search has and has not ruled out.
5. **Run one fresh confirmation** at those two values, with $B = 100000$ and a new recorded
   seed. The extra repetitions buy precision at those two sizes; they do not change which
   sizes you chose.
6. **State what it supports.** Read each row's `assessment`. If the candidate is `meets`, it
   is supported at that tested size; if the smaller one is `below`, it falls short; if
   either is `unresolved`, say so. Then say whether more repetitions at the same sizes, or
   more sizes, is what would resolve what is left.
7. **Put it in context.** Compare your finding with 25,550, and answer: what is the
   difference between a sufficient mathematical bound and a sample size supported by
   simulation? Does testing a grid of sizes prove the **smallest possible** $N$? Does a
   guarantee about the **endpoint** of an experiment guarantee that its entire **path**
   stays inside the band? Use your Part 2a plot on that last one.

::: callout-important
## How to read the simulation uncertainty

The helpers report `lower95` and `upper95`, a 95% Monte Carlo interval for $q_N$. It
reflects the finite number $B$ of experiments you ran. **You do not need to derive it.** It
describes the precision of your simulation estimate, and it is a separate thing from your
chosen $1000:1$ requirement.

| `assessment` | Interpretation at this tested $N$ |
|---|---|
| `meets` | The whole interval is at or above $c/(c+1)$ |
| `below` | The whole interval is below $c/(c+1)$ |
| `unresolved` | The interval straddles the target — this simulation is not precise enough to decide |

**`unresolved` is an acceptable answer.** Report it and say what further simulation would be
needed. What is *not* acceptable is rerunning with new seeds until a candidate passes: that
turns a confirmation into a search, and the fresh study stops meaning anything.
:::

**Answer in your `hw-02.qmd`:** the pilot and refined tables, the probability plot, and the
fresh confirmation table. Summarize the pilot and the refinement; name and justify your
candidate and your comparison; state your conclusion from the confirmation; answer all
three questions in Question 7.

::: render-commit-push
Commit and push now: *"finished problem 2."*
:::

# Problem 3 — Standards for two honest games

## Part 3a — Define two events and standards

You will study **both** games below. For each one, the honest model must specify the
randomization and have a success chance $p_0$ **that you can calculate by counting**. A
trial may contain several actions; different trials must start afresh and be independent.

| Study | One trial | A success | Counting hint |
|---|---|---|---|
| **Heads** | Toss a fair coin four times | All four tosses are heads | Count equally likely ordered sequences |
| **Flush** | Deal five cards from a freshly shuffled standard 52-card deck, no jokers | All five cards share a suit | Count five-card subsets with `choose()` |

::: callout-warning
## The trap in each one

**Heads** — use separate four-toss blocks. Overlapping windows in one long sequence are
**not** independent trials.

**Flush** — draw without replacement *within* a hand, then return all five cards and
reshuffle before the next hand. Count straight flushes and royal flushes as successes too:
the event here is simply "all one suit."
:::

Before you run any large simulation, write down — **separately for each of the two
studies**:

1. Your event, what one trial means, your independence assumptions, and a **counting
   derivation** of $p_0$.
2. Your tolerance $\epsilon$ — the largest absolute difference between an observed success
   proportion and $p_0$ that you would accept. Use $0 < \epsilon < p_0$. A useful starting
   point is $\epsilon = p_0/5$, a 20% relative discrepancy.
3. Your certainty odds $c:1$. Possible choices are $19:1$, $99:1$, or $1000:1$; any
   $c \geq 9$ is fine if you justify it. State the required probability $c/(c+1)$.
4. A **concrete setting** in which you would use this standard, and why that tolerance and
   that certainty are reasonable there.



5. Then: explain what would go wrong if the flush study automatically reused $\epsilon = 1/50$.
Would seeing **no flushes at all** produce a proportion inside that band? How useful would
that be as a check on a card dealer?

**Answer in your `hw-02.qmd`:** both study specifications and the flush comparison. **Make
these choices before you go looking for the $N$ values that work** — choosing the standard
after seeing which $N$ passes is how a check stops being a check.

## Part 3b — Code the mechanisms before using the shortcut

Write `play_heads(N)` and `play_flush(N)`, each returning `N` success indicators. Use
`sample()` to generate the actual tosses and cards. **Do not use `rbinom()` in this part** —
the point is that you can build a mechanism from its description, and only then replace it
with a shortcut you have earned.

Your template gives a loop shell for each and syntax hints. Useful counting syntax:

```{r}
#| label: counting-syntax
choose(52, 5)
choose(13, 5)
```

Then, **for each study**:

1. Check the length and the possible values of the output — `check_indicators()` does both.
2. Simulate **50,000 trials** and compare their success proportion to your calculated $p_0$.
3. Explain any discrepancy, **including why a rare event can have noticeable relative
   simulation error** even when its absolute error is tiny.
4. Identify the line of code that makes successive trials start afresh.

::: callout-tip
## Put the two discrepancies next to each other

Question 3 is much easier to answer once both numbers are on the page. Run both simulators,
then write the two differences side by side — as absolute differences *and* as percentages
of their own $p_0$. One of those two columns will look like the simulators are equally good
and the other will not, and explaining why is the whole question.
:::

**Answer in your `hw-02.qmd`:** both functions, the output checks, both numerical
comparisons, and your explanation.

## Part 3c — Find N for both studies

Run the whole Part 2c workflow **twice, once per study**. Keep $p_0$, $\epsilon$ and $c$
matched to the event you are studying — a good heads standard is not automatically a good
flush standard, and mixing them is the easiest way to produce a table that means nothing.

| Stage | $B$ |
|---|---:|
| Exploration and refinement | 10,000 |
| Fresh confirmation | 100,000 |

For **each** study, in this order:

1. **Run the pilot grid** with `simulate_grid()` across at least six values of $N$. Start
   with the supplied grid and extend it if every tested value comes back `below`, or every
   one comes back `meets`.
2. **Read the table and the plot.** Find where `q_hat` and `assessment` move toward
   $c/(c+1)$. Keep the counts and the uncertainty bounds visible — **do not round `q_hat`
   to 1** before comparing it to the target.
3. **Refine near the transition.** Add at least four new values of $N$ around the promising
   or unresolved region. **Keep every value you tried**; do not replace earlier results with
   whichever run looks best.
4. **Choose before you confirm.** Pick one `candidate_N` and one smaller `comparison_N`, and
   record why those two.
5. **Run one fresh confirmation** at those two values with a new recorded seed and
   $B = 100000$.
6. **Report the smallest tested value** your confirmation supports, or an unresolved range.

**Do the heads study end to end first, then the flush study end to end.** Your submitted
tables must show **both the inside/outside counts and the Monte Carlo uncertainty**, not
just a rounded success percentage.

::: callout-important
## Do not claim an exact universal cutoff

Simulations need not give a perfectly increasing curve, and the possible success counts are
integers. For the flush in particular, check what whole numbers of successes your band
actually admits at small $N$ before you conclude anything from a low `q_hat` there — the
answer is occasionally *none*, and a band no outcome can land in is a property of your
standard, not a fact about dealers.
:::


7. Finally, **put the two studies side by side** in the table your template provides: $p_0$,
$\epsilon$, your confirmed $N$, and the expected number of successes at that $N$. Same rule,
two events — **what did changing only the rarity of the event cost you?**


**Answer in your `hw-02.qmd`:** the search and confirmation tables for both studies, a plot
per study with the target marked, your recommendation for $N$ in each, and comparison of two experiments. 

## Part 3d — What would make you trust it?

::: callout-note
**Discuss it — do not compute anything.** Everything you need is the $\epsilon$ you chose in
Part 3a and what Part 2a showed about where an observed proportion settles.
:::

Part 3c took $p_0$ as known and asked how many trials make an honest mechanism's observed
proportion land near it. Suppose instead the mechanism you are auditing was never at $p_0$
— its true chance is a little higher, and nobody told you.

Would your check object? Would running more trials make it **more** likely to object, or
**less**? If the difference is small enough that your check will never object at any number
of trials, is that a defect in the procedure or the thing you asked for in Part 3a? Would it
matter in the setting you named there?

**Answer in your `hw-02.qmd`:** one paragraph.

::: render-commit-push
Commit and push once all three problems are done, with a message describing what's in the
commit.
:::

# Submission

::: callout-warning
Before you wrap up, make sure your Git pane is empty — everything committed and pushed.

You must also **render your document to PDF and submit that PDF on Gradescope** before the
deadline for full credit.
:::

Your submitted document must contain the probability work, all requested code and tables,
the four main plots (the urn paths, the Bernoulli search, and the two game searches), and
your Part 3d paragraph.
**Record your seeds** and label $N$, $B$, $p_0$, $\epsilon$, and $c$ so that someone else
can reproduce your study. No separate essay is required.

To submit on Gradescope:

- Access Gradescope through the menu on the
  [STA 119FS Canvas site](https://canvas.duke.edu/courses/84980).
- Click on the assignment, and you'll be prompted to submit it.
- Mark the pages associated with each problem — every page of your submission should be
  associated with at least one question.
