---
title: "Lab 01: Mayer's method of averages"
subtitle: "27 equations, 3 unknowns, and no exact answer"
execute:
  freeze: auto
  echo: true
  eval: false
  warning: false
  message: false
format:
  html:
    toc: true
    code-copy: true
---

::: callout-important
## Due date

This lab is due **today, by the end of class**, per this course's usual rhythm — everyone
commits and pushes before the exit ticket. There is nothing to submit outside your repo;
what's on GitHub when class ends is what's graded (completion-based — see Grading below).
:::

# Introduction

In 1750 Tobias Mayer wanted to measure the libration of the moon — the slight wobble that
lets us see a bit around its edge. He took 27 observations, and each one gave him an
equation in the same 3 unknowns.

**27 equations, 3 unknowns.** More information than he needed. That sounds like good news.

It wasn't, and here's why: the 27 equations **contradict each other.** There is no set of
three numbers that satisfies all 27. There isn't even a set that satisfies four of them.
Every observation has a little error in it, so every equation is slightly wrong, and
they're all slightly wrong in different directions.

At the same time, Leonhard Euler — a far greater mathematician than Mayer — had 75
equations in 8 unknowns for the orbits of Jupiter and Saturn. **He gave up.** He didn't
believe you were allowed to combine measurements that disagreed; he thought errors would
pile up rather than cancel.

Mayer wasn't a better mathematician. He just wasn't afraid of the data. Today you find out
what he did.

::: callout-note
This is the first lab where you'll actually write and run R. `lab-00` only got
your tools working — there was no statistics in it on purpose. Today there is.
:::

## Learning goals

By the end of this lab you will be able to...

- pull rows and columns out of a data frame to build exactly the system of equations you want
- use `solve()` to find the exact solution to a square linear system, and explain *why* it
  fails on a non-square one
- explain, from having watched it happen on the board, why the answer to an overdetermined
  system depends on which rows you use
- reproduce Mayer's grouped-sum method and evaluate whether his grouping rule is doing real
  work or just decoration
- commit and push your work as the record of what you did, including what you asked an AI
  model and what it got wrong

---

# Getting started

## Clone the repo and start a new RStudio project

This is your **first private repo** for this course — not the public one from
`lab-00`. Your SSH key from that lab still works; you won't set it up again.

- Go to the invitation link posted for **Lab 01** and accept it. GitHub Classroom will
  create a private repo just for you, named something like `lab-01-yourusername`.
- Open your container and go to the repo's GitHub page. Click the green **Code** button,
  make sure **SSH** is selected (the label should read *Clone with SSH*, and the address
  should start with `git@github.com:`, not `https://`), and copy it.
- In RStudio: **File → New Project → Version Control → Git**, paste the address into
  **Repository URL**, leave the directory name as it fills in, and click **Create Project**.
- RStudio will clone the repo and open it. You should see `instructions.qmd`,
  `answers.qmd`, and a `data/` folder in the **Files** pane.
- Open `answers.qmd`. That is the file you actually type into today — **leave this file,
  `instructions.qmd`, alone.**

::: callout-tip
## If step one gives you an `https://` address instead

Switch it to SSH before you clone. An `https://` address will ask for a password on every
single push and will not accept your GitHub password — this is the single most common way
this step fails, and it's the same warning `lab-00` gave you.
:::

## Committing and pushing, for real this time

`lab-00` cloned a repo and stopped there — no committing, no pushing. Today you
do both, more than once.

**To commit:** in the **Git** pane (top right), check the box next to `answers.qmd` to
stage it, write a short message describing what changed (e.g. "Step 1 and 2 answers"), and
click **Commit**.

**To push:** click **Push**. This sends your commit to GitHub. Refresh your repo's GitHub
page and you should see the new commit there.

You'll be asked to commit and push at several points in this lab — treat each one as a
checkpoint, not a formality. **Two of those commits matter more than the rest**: one right
before you use any AI assistance in Section 5, and one right after. The diff between them
*is* this course's AI-disclosure mechanism — see `AI_SPINE.md` if you want the full
reasoning.

---

# Packages

None today. This course starts in base R on purpose — `solve()`, `[ , ]` subsetting, and
`data.frame` indexing are all you need, and adding a package before you've felt what base R
does would bury the actual skill under syntax. `ggplot2` arrives around Lab 5–6, once
layered plots are genuinely useful.

```{r}
#| label: load-data
d <- read.csv("data/mayer_equations.csv")
dim(d)
head(d)
```

27 rows. Each row is one equation: `x1_coef * x1 + x2_coef * x2 + x3_coef * x3 = rhs`.

---

# Exercises

**Goal:** turn 27 contradictory equations into a single, defensible claim about 3 unknown
quantities — and understand why that turn was Mayer's invention, not the arithmetic after
it.

## 1. Predict (5 minutes)

Write both of these in `answers.qmd` **before you touch R.**

> **Prediction 1.** You're going to pick 3 of the 27 equations and solve them exactly. Then
> another group picks a *different* 3. How close will the two answers be?

> **Prediction 2.** Mayer added nine of his equations together to get one equation. He
> claimed this made his answer **nine times** more accurate than a single observation. Was
> he right? If not, what's the right number?

We'll take a show of hands on the second one. **The answer doesn't come until Exercise 4.**

## 2. Pick three equations and solve them (13 minutes)

```{r}
#| label: solve-three
rows <- c(1, 2, 3)

A <- as.matrix(d[rows, c("x1_coef", "x2_coef", "x3_coef")])
b <- d$rhs[rows]

A          # look at it. 3 rows, 3 columns.
solve(A, b)
```

Three numbers out. That's an *exact* answer to those three equations.

Now try it with all 27:

```{r}
#| label: solve-all-27
A_all <- as.matrix(d[, c("x1_coef", "x2_coef", "x3_coef")])
solve(A_all, d$rhs)
```

::: callout-important
## Read that error message

R is not being difficult. It is telling you something true about the world: **there is no
exact answer, and it cannot pretend otherwise.**

Mayer stared at this same wall in 1750, without R to tell him about it.
:::

**New tool today:** `solve()`. The real skill is `d[rows, cols]` — pulling out the rows you
want.

::: render-commit-push
Commit and push now, with a message like "loaded data, solved a 3x3." Everyone in your
group does this on their own machine — this course doesn't use a driver/navigator split;
you all type everything.
:::

## 3. Group build (28 minutes)

### Step 1 — recreate the crisis (10 min)

Pick **any three** of the 27 equations. Not 1, 2, 3 — pick three you like. Solve them.

Put your three numbers on the board under your group letter. Now look at the board.

Then do it again with three *different* rows, in your own group. Put that up too.

::: callout-tip
## This is the whole of Chapter 1

You can get almost any answer you want out of this data, depending on which rows you
happen to grab. Every single one of those answers is *exactly* correct — for the three
equations you chose.

Now you understand why 18th-century astronomers were nervous.
:::

### Step 2 — Mayer's fix (10 min)

Mayer's idea: don't throw 24 equations away. **Add them up.**

Sort the 27 equations into three groups of nine, add each group together to make one
equation, and solve the resulting 3×3.

His rule for *how* to sort them: put equations in different groups when their coefficients
are as different as possible. Note that this is a rule about the *coefficients*, not about
the answers — he is choosing his groups before he knows what they'll give him.

```{r}
#| label: mayer-grouping
# sort by the second coefficient, then take thirds
o <- order(d$x2_coef)
g1 <- o[1:9]; g2 <- o[10:18]; g3 <- o[19:27]

sum_rows <- function(idx) {
  c(colSums(d[idx, c("x1_coef","x2_coef","x3_coef")]), rhs = sum(d$rhs[idx]))
}

M <- rbind(sum_rows(g1), sum_rows(g2), sum_rows(g3))
M

A <- M[, 1:3]
b <- M[, "rhs"]
solve(A, b)
```

**One answer.** Put it on the board.

Now test whether his sorting rule actually matters. Group them a different way — just rows
1–9, 10–18, 19–27, in whatever order the file happens to be in:

```{r}
#| label: file-order-grouping
g1 <- 1:9; g2 <- 10:18; g3 <- 19:27
# ... same three lines again
```

**Did the answer move? More or less than it moved in Step 1?**

::: callout-note
## Careful what you conclude from that

Sorting by `x2_coef` is not Mayer's rule — it's a **proxy** for it. His rule is *spread the
coefficients out between groups*, and sorting on one column can quietly put rows with very
different values of the **other** column into the same group, destroying exactly the
contrast you need to pin that coefficient down.

So the sorted grouping is **not guaranteed to win**, and if yours doesn't, that is a
finding, not a mistake. Report what you got.

The comparison that *is* robust is the one against Step 1: two groupings disagree with each
other far less than two three-row picks did. **Which tells you where the accuracy came
from — the summing or the sorting?**
:::

### Step 3 — the modern tool (8 min)

```{r}
#| label: lm-comparison
fit <- lm(rhs ~ x2_coef + x3_coef, data = d)
coef(fit)
```

All 27 rows, no grouping, no choosing. A different answer again.

It is better than Mayer's, in a specific sense — but that sense doesn't exist yet in this
book. It gets invented in Chapter 4. **Hold the question.**

::: render-commit-push
Commit and push: "Steps 1-3 complete." This is a good moment for that Git pane to go
empty before the debrief.
:::

## 4. Debrief (12 minutes)

**Closing question:** Your group got one answer from three equations. Mayer got one from
all 27 by adding them up. `lm()` gets a third. **Which one goes in the paper — and how do
you defend it to a colleague who picked three different rows than you did?**

Write your position in `answers.qmd`, and the strongest objection to it.

Also be ready to discuss:

- Euler had the same problem, better mathematics, and quit. **What did Mayer have that
  Euler didn't?**
- Mayer's rule says to spread the coefficients out across groups. Why would that help?
  (You are 170 years early to the field called experimental design.)

### Prediction 2, resolved

Mayer summed nine equations and claimed his answer was **nine** times more accurate than a
single observation.

It's **three**. $\sqrt{9}$.

He was wrong, in print, in the founding example of the subject. Remember this number — it
comes back in every lab for the rest of the semester.

**Is that the kind of mistake that invalidates his work?** Argue about it.

## 5. AI slot (5 minutes)

::: callout-important
**Commit first — before you type anything into an AI tool.** The commit you're about to
make is the "before" half of this course's disclosure record.
:::

Ask your model exactly this:

> *Write R code implementing Tobias Mayer's method of averages to solve an overdetermined
> system of 27 linear equations in 3 unknowns.*

You have just spent half an hour doing Mayer's method by hand. So you're in an unusually
good position to answer one question, which is the only thing you have to write down:

**Did it do Mayer's method, or did it do least squares with Mayer's name on it?**

Look specifically for the grouping rule — sorting into three groups so the coefficients
differ as much as possible. That rule *is* the invention. If it's missing, whatever you got
back isn't the method of averages, however confidently it's labelled.

::: callout-note
## Why this keeps happening

**This is an easy course about hard-to-find methods.** These models are excellent at the
thing everyone does and unreliable at the thing one person did in 1750, once, and nobody
has done since.

You will see this specific failure about a dozen more times this semester. Start keeping
score.
:::

::: render-commit-push
**Commit again** — this is the "after" half of the disclosure record. Push.
:::

## 6. Exit ticket (5 minutes)

On paper, handed in as you leave — not in `answers.qmd`:

1. What number did your group get, and how did it compare to your prediction?
2. Muddiest point?
3. One thing you'd want to try in R next.
4. What did you personally do today?

---

# Submission

Nothing goes to Gradescope in this course — your GitHub repo **is** the submission.

::: callout-warning
Before you leave, make sure `answers.qmd` is fully committed and pushed. Check your Git
pane: it should be **empty**. If it isn't, something didn't push, and that's the thing to
fix before the exit ticket, not after.
:::

To confirm you're done:

- Open your repo on GitHub and refresh the page.
- You should see **at least two commits** — one from before Section 5's AI slot, one from
  after.
- `answers.qmd` on GitHub should show your predictions, Steps 1–3, the closing-question
  position and objection, and the AI-slot answer.

# Grading

This course grades labs on **completion, not correctness** (see `course_design.md` §5) —
**2** = every box below checked, **1** = partial, **0** = nothing pushed.

| Check | What it means |
|---|---|
| Every code chunk attempted | Running with an error still counts — `error: true` is set in `answers.qmd` so a broken chunk still renders |
| Both predictions recorded | Before any computation happened |
| Steps 1–3 all recorded | Both row-choice answers from Step 1, Mayer's grouped answer, the file-order comparison, and the `lm()` answer |
| Closing-question position | 2+ sentences, plus the strongest objection to it |
| One error and how you handled it | Doesn't have to be a big one |
| Two commits, pushed | Before the AI slot, and after |

A wrong final number with clearly shown reasoning earns full credit. A suspiciously clean
number with no visible work does not.
