---
title: "Lab 03: How many trials are enough?"
execute:
  freeze: auto
  echo: true
  warning: false
  message: false
format:
  html:
    toc: true
    toc-depth: 3
    code-copy: true
editor: 
  markdown: 
    wrap: 72
---

::: callout-important
## Due date

This lab is due **today, by the end of class**. To be considered on
time:

- `lab-03.qmd` committed and pushed to your GitHub repo before the exit
  ticket
- That same document, rendered to PDF, submitted on Gradescope
:::

# Introduction

Consider the following experiment: **roll a die four times, and win if
at least one six turns up.** Today you will build this experiment in R,
and then ask a question: **how many times would you have to play before
the dice told you anything?**

## Learning goals

By the end of this lab you will be able to...

- **count a probability through its complement**.
- write a simulator for an event whose **one trial contains several
  actions**.
- understand the difference between $N$ (the trials in one experiment)
  and $B$ (repetitions of the whole experiment).
- state what passing a frequency check does and does not establish about
  a die.

## Getting your repo

**1 · Find the repo.** The
[statistical-history](https://github.com/statistical-history)
organization, the repo whose name starts with `lab-03`.

**2 · Clone it.** RStudio: **File → New Project → Version Control →
Git**, paste the SSH address, **Create Project**. You should see
`lab-03.qmd` and `lab-03-helpers.R`.

**3 · Names in the YAML**, then **Render** once before you change
anything.

::: callout-tip
## Pull first. Yes, again.

It is your own repo and nobody else has touched it. Pull anyway — the
habit is the point, and the one week you skip it is the week it matters.
:::

# Part 1 — Make a guess.

## Before anyone opens a laptop

**Write your guess.** No computing and no looking anything up. The only
wrong answer is a blank one.

> **Guess.** Someone hands you a die and says it is fair. You are
> allowed to roll it in blocks of four and count how often at least one
> six turns up. **How many blocks would you want before you would be
> willing to say the die looks honest?** One number.

# Part 2 — Count it first

**Goal:** get the honest probability by counting.

One trial is **four rolls of a fair die**, and a trial is a **success**
if at least one of the four is a six.

The direct count is a nuisance: you would have to add up the cases of
exactly one six, exactly two, exactly three, and exactly four, and it is
very easy to double-count. **Count the complement instead** — the trials
with *no* six at all.

Four rolls give $6^4 = 1296$ equally likely ordered outcomes. A trial
with no six uses only the other five faces, so there are $5^4 = 625$ of
those. Everything else has at least one six:

$$p_0 = 1 - \frac{5^4}{6^4} = 1 - \frac{625}{1296} = \frac{671}{1296}.$$

Fill in a code chunk in your template to calculate and save this
probability.

```{r}
#| label: count-six-probability-template
#| eval: false
p0_six <-NA # replace with your code
p0_six
```

```{r}
#| label: count-six-probability
#| include: false
p0_six <- 1 - (5 / 6)^4
p0_six
```

::: callout-tip
## Stop and answer — Part 2

1.  Why is this **not** $1/6$? And why is it not $4 \times 1/6 = 2/3$?
    Say what each of those two wrong answers is actually the probability
    *of*.
2.  Suppose the game were **six** rolls instead of four. Write the
    probability down the same way and compute it. Which direction did it
    move, and does the complement trick still work?
:::

# Part 3 — Build the mechanism

```{r}
#| label: setup

options(digits = 7)

source("lab-03-helpers.R")
```

::: callout-warning
## Three of today's functions are not R

`check_indicators()`, `simulate_grid()` and `plot_certainty()` were
written for this course. They live in `lab-03-helpers.R`, which is the
file you just sourced. **Open the file** if you want to see what each
function does in more details; there is an index at the top and comments
above each function.

| Call | What it does for you | First used |
|------------------------|------------------------|------------------------|
| `check_indicators(x, N)` | Confirms a vector is exactly `N` zeros and ones. Prints nothing when happy | Part 3 |
| `simulate_grid(N_values, p0, epsilon, certainty_odds, B, model_p)` | Runs a whole study at each `N` you name and stacks the results into one table | Part 5 |
| `plot_certainty(results, certainty_odds)` | Plots one of those tables, with the target marked | Part 6 |

There is a fourth, `summarize_experiments()`, which turns a pile of
success counts into one row of that table. **You never call it** —
`simulate_grid()` calls it for you, once per `N`. It is in the file if
you are curious.

Everything else today is **real base R** and is documented, so `?sample`
works: `sample()`, `rbinom()`, `numeric()`, `seq_len()`, `any()`,
`as.integer()`, `mean()`, `table()`, `data.frame()` and `set.seed()`.
:::

**Goal:** a function that plays the game `N` times and reports, for each
of those `N` trials, whether it was a success.

```{r}
#| label: play-six-function
play_six <- function(N) {
  outcomes <- numeric(N)

  for (i in seq_len(N)) {
    rolls <- sample(1:6, size = 4, replace = TRUE)
    outcomes[i] <- as.integer(any(rolls == 6))
  }

  outcomes
}
```

Line by line:

| The code | What it does |
|------------------------------------|------------------------------------|
| `numeric(N)` | Makes room for `N` results before the loop starts |
| `seq_len(N)` | The trial numbers, 1 through `N` |
| `sample(1:6, size = 4, replace = TRUE)` | Rolls the die **four times** — one whole trial |
| `replace = TRUE` | Rolling a 3 does not remove the 3 from the die |
| `any(rolls == 6)` | `TRUE` if at least one of the four was a six |
| `as.integer()` | Turns `TRUE`/`FALSE` into `1`/`0` so it can be averaged |

```{r}
#| label: check-six-function
set.seed(20260918)

small_run <- play_six(20)
check_indicators(small_run, 20)
table(small_run)

large_run <- play_six(50000)
check_indicators(large_run, 50000)

data.frame(
  calculated_p0 = p0_six,
  simulated_rate = mean(large_run),
  difference = mean(large_run) - p0_six
)
```

`check_indicators()` prints **nothing** when it is happy. That is the
intended output. It stops with an error if the vector is the wrong
length or contains something other than 0 and 1 — and note what it does
*not* check: it has no idea whether you rolled a six-sided die or a
twenty-sided one.

::: callout-note
## `set.seed()` — what it does, and where it must not go

R's randomness is not really random; it is a long fixed sequence, and
`set.seed(n)` says *start reading at position `n`.* Two people who set
the same seed get the same rolls, which is what makes a simulation
reproducible — and why every chunk today carries one.

**Set it before a study, never inside a function.** If `set.seed()` sat
inside `play_six()`, every trial would restart the sequence at the same
place and all `N` trials would come out identical. The seed above is in
the chunk, outside the function, which is exactly right.
:::

::: callout-tip
## Stop and answer — Part 3

1.  **Which line makes a new trial start afresh (runs a new
    experiment of rolling a die 4 times)?** Point at it.
2.  Suppose instead you rolled the die 200 times in one long sequence
    and slid a four-roll window along it — rolls 1–4, then 2–5, then
    3–6. That also gives you "trials." **Why are those not
    independent?**
3.  What does `calculated_p0` tell us and what does `simulated_rate`
    tell us? How are the two different?
:::

# Part 4 — What we are actually going to do

**Goal:** before choosing any numbers, get the whole procedure straight.
Everything after this is just running it.

Part 3 gave you one experiment, `play_six(N)`, that plays the game `N`
times and `mean()` turns successes of each game into a single
proportion. The question this lab exists to answer is what should `N` be
and what to *do* with that number when someone hands you a die and says
it is fair.

Here is the procedure, start to finish.

**· The claim.** Someone says the die is fair. **If they are right**,
one trial — four rolls, looking for at least one six — succeeds with
chance $p_0 = 0.5177$. That is the number you counted in Part 2, and it
is the only thing the claim buys us.

**· What "chance** $0.5177$" actually promises. Not that an experiment
with a fixed `N` will give you $0.5177$. It promises that as you keep
playing, the proportion of trials that succeed settles near $0.5177$.
You already saw this in Part 3: 50,000 trials gave **0.5153**, close to
the counted value but not equal to it.

**· One experiment gives one number.** Play $N$ trials, count $m$
successes, and you have $\widehat p = m/N$. **Run it again and you get a
different one** — different rolls, different count. That is the whole
difficulty. You are trying to judge a fixed die using a number that
moves.

**· So we allow a tolerance.** How far may $\widehat p$ sit from $p_0$
before we object? Call that distance $\epsilon$. The interval

$$[\,p_0 - \epsilon,\; p_0 + \epsilon\,]$$

is the **acceptance band**, and one experiment **passes the check** if
its $\widehat p$ lands inside it. That phrase — *passes the check* — is
what the rest of the lab is about.

**· And we set a standard.** Because $\widehat p$ moves, an honest die
will sometimes miss the band by bad luck. We want that to be rare. So we demand: **an honest die must
pass at least** $c/(c+1)$ of the time.

**· What we are actually looking for.** Every number here is computed
**assuming the die is fair** — $p_0 = 0.5177$ is what a fair die gives,
and the band is drawn around it. So the goal, stated as something we can
aim at:

> **Find an** $N$ big enough that a fair die almost never lands outside
> the band.

Turn that around and you have the reason anyone would want such an $N$.
If at that size a fair die falls outside only once in a hundred tries,
then **someone who runs a single experiment and lands outside has just
seen something that rarely happens to fair dice.** That is worth
flagging. It is a result you would not shrug at.

That is the whole use of $N$: **the experiment size at which landing
outside the band stops being ordinary and starts being worth a second
look.** Too small an $N$ and honest dice fall outside constantly, so the
flag means nothing.

We find it by simulation, because there we set the truth ourselves. Play
the whole experiment $B$ times **from a die we have deliberately made
fair**, and count how many passed. That proportion estimates

$$q_N = P\big(\,|\widehat p - p_0| \le \epsilon\,\big),$$

and we want $q_N \ge c/(c+1)$ — equivalently, **the chance of landing
outside is at most** $1/(c+1)$, which for $c = 99$ is one in a hundred.

::: callout-warning
## A threshold for surprise, not a verdict
Landing outside the band is **unlikely if the die is fair.** That is a
statement about fair dice. It is **not** the same as "the die is
probably unfair," and the two come apart in a way that matters. **The debrief pulls them
apart.**

For now: you are building a threshold for *surprise*, not a verdict.
:::

::: callout-important
## Two numbers, two jobs — this is the distinction the whole lab turns on

|   | What it is | What it controls |
|------------------------|------------------------|------------------------|
| $N$ | trials inside **one** experiment | how much $\widehat p$ bounces around. Bigger $N$, steadier $\widehat p$ |
| $B$ | how many **whole experiments** we simulate | how precisely we know $q_N$. Bigger $B$, sharper estimate |

$N$ is a fact about the *study someone would actually run*. $B$ exists
only because we are using a simulation to determine $N$. 
:::

## Now the two choices

Neither has a correct answer. Both are ours to make, and the debrief comes
back to how much they were doing.

| Choice | What it means | Ours in this Lab |
|------------------------|------------------------|------------------------|
| **Tolerance** $\epsilon$ | How far $\widehat p$ may sit from $p_0$ and still pass | $p_0/5$ — a 20% relative discrepancy |
| **Certainty odds** $c:1$ | How often an honest die must pass | $99:1$, i.e. probability $99/100$ |

```{r}
#| label: six-standard
epsilon_six <- p0_six / 5
c_odds_six <- 99
target_six <- c_odds_six / (c_odds_six + 1)

data.frame(
  p0 = p0_six,
  epsilon = epsilon_six,
  band_lower = p0_six - epsilon_six,
  band_upper = p0_six + epsilon_six,
  certainty_odds = c_odds_six,
  target_probability = target_six
)
```

`band_lower` and `band_upper` are the two ends of the acceptance band.
**An experiment passes if its** $\widehat p$ falls between them,
endpoints included.

::: callout-important
## Making $\epsilon$ a *fraction* of $p_0$ is itself a decision

We could have picked a flat number — 0.02, say, the way Bernoulli did
for his urn. Writing it as $p_0/5$ instead means the tolerance **scales
with the event**: a common event gets a wide band and a rare one gets a
narrow band.

That choice is doing more work than it looks like, and HW 02 makes you
feel it. Hold on to this one; the debrief comes back to it.
:::

::: callout-tip
## Stop and answer — Part 4

1.  Our band is $[0.414, 0.621]$ — **more than a fifth of the whole
    probability scale wide.** Name a use for this check where that width
    is fine, and one where it plainly is not.
2.  A die is loaded so that a six comes up with chance $1/5$ instead of
    $1/6$. That makes this event's probability $1 - (4/5)^4$. 
    Add a code chunk to calculate this value. Does this value fall inside the band we chose?
:::

# Part 5 — Run it once, for a fixed $N$.

**Goal:** see what the helper's table actually reports, at a large $N$.

Part 3 ran 50,000 trials. Let us keep that size, and run the
honest-model study described above: simulate $B = 10{,}000$ whole
experiments of 50,000 trials each, and count how many pass.

```{r}
#| label: six-fixed-N
set.seed(61000)

six_fixed <- simulate_grid(
  N_values = 50000,
  p0 = p0_six,
  epsilon = epsilon_six,
  certainty_odds = c_odds_six,
  B = 10000
)

six_fixed
```

All 10,000 experiments passed. At this size $\widehat p$ has essentially nowhere to go, so
that is no surprise.

| Column | What it is claiming | Here |
|---|---|---|
| `inside`, `outside` | How many of the $B$ experiments passed and failed. They add to $B$ | 10,000 and 0 |
| `q_hat` | `inside / B` — the estimated chance that a whole experiment passes | 1.0000 |
| `odds_hat` | `inside / outside`. **`NA` means no failures were seen**, not that the odds are infinite | `NA` |
| `lower95`, `upper95` | How much to trust `q_hat`, given how many experiments you ran — see below | 0.99963 to 1 |
| `assessment` | Those two, compared against the $99/100$ target | `meets` |

## How to read the simulation uncertainty

The helpers report `lower95` and `upper95`, a 95% Monte Carlo interval for $q_N$. You do not need to understand how it is built and what it means beyond that it describes
the precision of your simulation estimate, and it is a separate thing from your chosen
$99:1$ requirement.

| `assessment` | Interpretation at this tested $N$ |
|---|---|
| `meets` | The whole interval is at or above $c/(c+1)$ |
| `below` | The whole interval is below $c/(c+1)$ |
| `unresolved` | The interval straddles the target — this simulation is not precise enough to decide |

**If `assessment` says `unresolved`, it means $B$ is not high enough to make a decision about
$q_N$.** Not that the die failed, and not that this $N$ is wrong — only that this many
experiments cannot tell which side of the target the truth is on. The fix is more experiments,
not more trials. 


::: callout-tip
## Stop and answer — Part 5

1.  `q_hat` came back as exactly 1. **How many of the 10,000 experiments had their
    $\widehat p$ land inside the band?** And what does that have to do with `odds_hat`
    printing `NA`?
2.  Imagine you have a slider that controls $\epsilon$, keeping everything else the same. Suppose you drag it towards zero, thus, making the band **narrower** around the same $p_0$. **Would `q_hat` go up or down as the slider for epsilon gets closer to zero?**
:::

# Part 6 — Can we get away with less?

**Goal:** $N = 50{,}000$ works, but it is 200,000 individual rolls and
nobody is doing that at a table. We will now test a variety of different values of $N$ and try to find the smallest experiment size that
still clears the standard.

::: callout-note
## What we are really searching for
Remember, we are trying to find $N$ such that if we run the experiment and calculate $\hat p$, it would land within $\epsilon$ of the truth $p_0$ with probability $\frac{c}{c+1}$.  
:::
We start by checking a wide variety of values of $N$ and assess them with $B=10,000$ repetitions.

```{r}
#| label: six-pilot
set.seed(61001)

six_N <- c(25, 50, 100, 150, 200, 250, 300, 400, 500, 750, 1000)

six_pilot <- simulate_grid(
  N_values = six_N,
  p0 = p0_six,
  epsilon = epsilon_six,
  certainty_odds = c_odds_six,
  B = 10000
)

six_pilot
```

We now plot estimated $\hat q_N$ for each $N$ we tested. The dashed horizontal line corresponds to our goal of $q_N = 0.99$. 

```{r}
#| label: six-pilot-plot
plot_certainty(six_pilot, certainty_odds = c_odds_six)
```

**Read down the `assessment` column, not the `q_hat` column.** Three
rows matter:

| $N$ | `q_hat` | `assessment` | What it tells you |
|------------------:|------------------:|-----------------|-----------------|
| 100 | 0.9657 | `below` | Too small. The whole interval is under 0.99 |
| 150 | 0.9889 | `unresolved` | Above 0.99 as a point estimate, **and still not callable** — the interval crosses the line |
| 200 | 0.9968 | `meets` | Clears it, interval and all |

So the transition is somewhere between 100 and 200. We have looked at 150 but did not use sufficiently large $B$ to draw conclusions. Everything from 250 up also passes, but those are
more expensive and tell us nothing new. Zoom in on the window that is
actually in doubt.

```{r}
#| label: six-refinement
set.seed(61002)

six_refined_N <- c(105, 115, 125, 135, 145, 155, 165, 175, 185, 195)

six_refined <- simulate_grid(
  N_values = six_refined_N,
  p0 = p0_six,
  epsilon = epsilon_six,
  certainty_odds = c_odds_six,
  B = 10000
)

six_refined
```

145 and lower are `below`. 155 and 165 are still `unresolved`; 175 and up all read `meets`. **Notice what
that does and does not establish.** It says 175 passed at this $B$, on
this seed. It does not say 174 fails, and it does not say 175 is the
smallest possible — we tested ten numbers out of a hundred.


::: callout-tip
## Stop and answer — Part 6

1.  In the pilot, between which two values of $N$ does `assessment` stop
    saying `below`?
2.  155 comes back `unresolved` at the same $B$ as before. **What one change would
    settle if $155$ is large enough?**
3.  $N$ and $B$ both make things "more accurate," but not the same
    thing. One of them sharpens our estimate of $\widehat p$; the other
    sharpens our estimate of $P(\widehat p \text{ lands in the band})$.
    **Which does which** ?
:::

# Part 7 — Confirm it

**Goal:** one fresh study, at a new seed and ten times the experiments, on the values the
search could not settle.

## Which values are worth re-testing, and why $B$ jumps

The refinement left **155 and 165 `unresolved`** and put the first `meets` at **175**. Those
three are the ones to confirm — and notice **it is not the `below` rows that are worth
re-running.** 105 through 145 came back with their whole intervals under the target; running
them again at ten times the cost would only say `below` in more decimal places.

**The undecided ones are different.** `unresolved` means the study could not tell which side
of the target the truth was on. Look harder and it might come back `meets` — **which would
hand you a smaller, cheaper answer than 175.** That is the entire reason to spend the extra
experiments.

And $B$ goes from 10,000 to 100,000 because our target is 0.99: we are trying to measure a
failure rate of about one in a hundred, and 10,000 experiments expect only about 100
failures. Ten times the experiments, ten times the resolution — **on $q_N$, not on the die.**

::: callout-important
## Fix your values *before* you run this

A confirmation is only worth something if you chose what to confirm beforehand. Write 155,
165 and 175 down. Then run the chunk **once**.

**Rerunning with new seeds until something passes is not a confirmation** — it is a search
wearing a confirmation's clothes, and it makes the number meaningless.
:::

```{r}
#| label: six-confirmation
set.seed(61003)

six_confirm_N <- c(155, 165, 175)

six_confirmation <- simulate_grid(
  N_values = six_confirm_N,
  p0 = p0_six,
  epsilon = epsilon_six,
  certainty_odds = c_odds_six,
  B = 100000
)

six_confirmation
```

**165 resolved.** At $B = 10{,}000$ it was undecided; at $B = 100{,}000$ it reads `meets`, at
$\widehat q = 0.99203$ with the whole interval above the target. **So the answer improves from
175 to 165** — the extra experiments bought a smaller $N$, which is exactly what we hoped for
and why the smaller values were worth testing at all.

**155 did not resolve.** It is still `unresolved` at ten times the effort, sitting at
$0.99020$ with its interval straddling 0.99.

::: callout-warning
## Not every `unresolved` is a shortage of experiments

165 was undecided because $B$ was too small — look harder and it settles. **155 is a
different animal.** Its true pass rate sits almost exactly on 0.99, so no affordable number
of experiments will ever put it cleanly on one side. More $B$ would not help.

**That is a real and reportable result**, not a failure. The honest statement is: *165 is the
smallest tested size we can support; 155 we cannot resolve, and would need far more than
100,000 experiments to try.*
:::

::: callout-tip
## Stop and answer — Part 7

1.  The refinement said 175 and the confirmation says 165. **Which is the better answer, and
    what did the extra experiments actually buy you?**
2.  105 through 145 all came back `below` in the refinement, and we did **not** re-run them
    here. **Why not?** What would testing them again have told you?
:::

# Part 8 — Debrief

**Back together as a room.**

## One — you found $N$. Now what?

::: callout-important
## The question Part 4 left open

Everything today assumed the die is fair. So the study cannot *tell* you
a die is fair — it only found how big an experiment a **fair** die
needs.

Now hand that $N$ to someone holding a real die they know nothing about.
They roll $N$ trials and get one $\widehat p$.
:::

Suppose their $\widehat p$ lands **outside** the band. You built the check so that a fair
die stays inside 99 times out of 100. **So what has just happened, and what should they
think?**

Say it carefully — and then say what they should think if it lands **inside** instead.

## Two — which way does $N$ move?

Everything today held $\epsilon$, $c$ and $p_0$ fixed and hunted for $N$. Now turn the
handle the other way: change one of those three and ask **what $N$ would have to do to keep
the same standard.**

**Say bigger, smaller, or no change — and why.**

**(a)** We make $\epsilon$ **smaller** — a narrower band, everything else the same.

**(b)** We make $c$ **larger** — say $999:1$ instead of $99:1$.

# Part 9 — AI slot: the Decision Audit

Open a fresh chat and give it exactly this, and nothing else:

> *I roll a die 7 times. What is the probability I get at least one
> six?*

That is Part 2's question with a 7 in place of the 4, so **you can check
its arithmetic yourself** — and it will almost certainly get the
arithmetic right. **The answer is not what we are auditing.**

Your deliverable is a list of **everything it had to decide before it
could answer at all**, none of which you told it.

 You can start here.

| It had to assume something about… | Ask yourself |
|------------------------------------|------------------------------------|
| **How many sides** the die has | You never said six. Did it say it was assuming six, or just use it? |
| **Whether the die is fair** | Did it ask? Did it flag it? |
| **Whether the rolls are independent** | How does it combine probabilities from different rolls|
| **What you wanted it for** | A homework answer, a bet, a dice-tower audit? The number is the same; what you should *do* with it is not |


**Hand in** a list of things it had to decide on and **which of those it told you about**.

## Exit ticket

1.  Muddiest point?
2.  One thing you've learned today.
3.  What is your home town?

Make sure your name is on it as this is how attendance is taken.
